A dynamic programming algorithm to assess the reliability of inferences from network enrichment analysis

### 2018-05-02 doc/genesetdp.tex
More typos

Here, we present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the score distribution of any query of a given size. The algorithm enables us to calculate well defined statistics of inferences from NEA.  Using simulations we demonstrated that the algorithm produced unbiased statistics.

@@

In network-based gene set analysis one score a query set of genes $ \{g_1 \ldots g_Q\} $ from a genome with genes $\{g_1 \ldots g_G\}$ ($Q \le G$), and how they relate to a pathway $\{p_1 \ldots p_P\}$. In NEA the pathway maps to the genome through a network ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise. The pathways are scored by summing up all conections, i.e. the pathway score is

@@

Here, we wanted to determine a score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$. This allows us to evaluate our score against the null hyptesis, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''. This can be expressed as a mid $p$~value\cite{lancaster1961significance,hwang2001optimality} for obtaining a score $s$,

@@

Such number-of-links, $l_i$, can be precomputed ahead of time for all genes $i, \ldots, G$.
Furthermore, the number of genes having $a$ links to the investigated pathway is given by $k_a=\sum_{\{i:l_i=a\}}1$, and $R=\max_{i}{l_i}$.

@@

We dowloaded network definitions and the BinoX software (on 2018-04-28) for comparisons from \url{https://bitbucket.org/sonnhammergroup/binox}. BinoX was run with the default parameters.
% FIXME: add a more detailed description of the test set (i.e. organism and number of genes, pathways and network connections).

We also downloaded the NEA example given at the BinoX web site. The example files include a pathway definition file that groups $6819$ genes into $289$ human pathways, and a network definition file that, after thresholding with a score of $0.7$, yeilds $1244992$ links between those genes.

As a representative example for we chose the Glycolysis/Gluconeogenesis pathway, as it contained a number of genes that coincided with the median number of genes of the pathway definition file.

@@

We implemented a Python program that read network and pathway definition files and scored query sets according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}. We downloaded pathway and network definitions from the BinoX's website and used the same network treshold as BinoX default ($0.7$).

@@

It should be noted that our outlined method a score distribution could work well with Monte Carlo-sampling as well, by sampling random gene sets of size $Q$, which would be faster and easier to understand then the theory of how to permute networks.

>>>>

Here, we present a dynamic programming algorithm, that we dubbed GeneSetDP, that calculates the score distribution of any query of a given size. The algorithm enables us to calculate well-defined statistics of inferences from NEA.  Using simulations we demonstrated that the algorithm produced unbiased statistics.

@@

In network-based gene set analysis one score a query set of genes $ \{g_1 \ldots g_Q\} $ from a genome with genes $\{g_1 \ldots g_G\}$ ($Q \le G$), and how they relate to a pathway $\{p_1 \ldots p_P\}$. In NEA the pathway maps to the genome through a network ${x_{ij}}$, where $x_{ij}=1$ if $g_i$ and $p_j$ are connected, or $x_{ij}=0$ otherwise. The pathways are scored by summing up all connections, i.e. the pathway score is

@@

Here, we wanted to determine a score distribution, {\em i.e.} the number of ways, $N(s)$, we can pick $Q$ genes and obtaining a score, $s$. This allows us to evaluate our score against the null hypothesis, $H_0$ : ``The Q query genes are randomly selected from the genes in the genome''. This can be expressed as a mid $p$~value\cite{lancaster1961significance,hwang2001optimality} for obtaining a score $s$,

@@

Such number-of-links, $l_i$, can be pre-computed for all genes $i, \ldots, G$.
Furthermore, the number of genes having $a$ links to the investigated pathway is given by $k_a=\sum_{\{i:l_i=a\}}1$, and $R=\max_{i}{l_i}$.

@@

We downloaded network definitions and the BinoX software (on 2018-04-28) for comparisons from \url{https://bitbucket.org/sonnhammergroup/binox}. BinoX was run with the default parameters.
% FIXME: add a more detailed description of the test set (i.e. organism and number of genes, pathways and network connections).

We also downloaded the NEA example given at the BinoX web site. The example files include a pathway definition file that groups $6819$ genes into $289$ human pathways and a network definition file that, after thresholding with a score of $0.7$, gave $1244992$ links between those genes.

As a representative example for we chose the ``Glycolysis/Gluconeogenesis'' pathway, as it contained a number of genes that coincided with the median number of genes of the pathway definition file.

@@

We implemented a Python program that read network and pathway definition files and scored query sets according to Equation \ref{eq:sum_i}, the Dynamic Programming algorithm described in the Algorithm section, that enabled us to assign $p$~values according to Equation \ref{eq:pval}. We downloaded pathway and network definitions from the BinoX's website and used the same network threshold as BinoX default ($0.7$).

@@

It should be noted that our outlined method a score distribution could work well with Monte Carlo-sampling as well, by sampling random gene sets of size $Q$, which would be faster and easier to understand than the theory of how to permute networks.
